---
title: "RDML Data Pre-Processing Module"
author: "Abinaya Yogasekaram"
date: "02/07/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## RDML Data Pre-Processing Module 

Using the formatting functions that will convert different qPCR machine outputs into a standardized file type that is compatible with analysis with MDMAPR.
This script follows the workflow below:

1. Input both the fluorescence data and metadata.
2. Ensure the order of the metadata matches the order of the fluorescence data
3. Remove any null records (controls), and merge both sets of data
4. Extract the fluorescence data from the available records 
5. Extract the supporting metadata (reactionID, well,Location, Etc.)
6. Organize and compress the data into an RDML file.

This Tutorial/Module was developed using the [vignette](https://mran.microsoft.com/snapshot/2017-07-30/web/packages/RDML/vignettes/RDML.html) provided by the RDML consortium. 

### Loading Required Packges

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Packages used
packages = c("tidyverse","RDML","readxl")
## Load or install&load
package.check <- lapply(packages,
                        FUN = function(x) {
                          if (!require(x, character.only = TRUE)) {
                            install.packages(x, dependencies = TRUE)
                            library(x, character.only = TRUE)
                          }
                        }
)

```

### Read-in Required Helper Functions

```{r}
# helper function to read in the metadata sheets as separate dataframes all contained within a list,

# ensure data type for fluorescence file, should be numeric and will convert to numeric if not
data_type_check <- function (raw_fluorescence) {
  verfied_fluorescence <- as.data.frame(apply(raw_fluorescence[,2:ncol(raw_fluorescence)], 2, function(x)
    as.numeric(as.character(x))),
    str(raw_fluorescence))
  verfied_fluorescence$wellLocation <- raw_fluorescence$wellLocation
  verfied_fluorescence <- as.data.frame(verfied_fluorescence[,c(ncol(verfied_fluorescence),1:ncol(verfied_fluorescence)-1)])
  print("data are all numeric")
  return(verfied_fluorescence)
}

# names of the list elements are the same as the sheet titles
read_excel_allsheets <- function(filename, tibble = FALSE) {
  sheets <- readxl::excel_sheets(filename)
  x <- lapply(sheets, function(X) readxl::read_excel(filename, sheet = X))
  if(!tibble) x <- lapply(x, as.data.frame)
  names(x) <- sheets
  x
}

# format the metadata excel file
format_qPCR_metadata_RDML <- function(metadataFile) {
  
  #Creating metadata table
  metadata_table <-
    distinct(left_join(metadataFile$assay_Table, metadataFile$standardCurveResults_Table[, c("standardCurveID", "assayID", "standardCurveName", "SCdate", "SCrecordedBy", "SCdataNotes")], by = "assayID"))
  
  metadata_table <-  left_join(metadataFile$results_Table, metadata_table, by = "assayID")
  
  metadata_table <- left_join(metadata_table, metadataFile$replicate_Table, by = "extractID")
  
  metadata_table <- left_join(metadata_table, metadataFile$project_Table, by = "stationID")
  
  
  #Organize columns
  metadata_table <- as.data.frame(metadata_table[, c(
    "resultID","runID", "assayID","pcrChemistryID","extractID","wellLocation","sampleName", "copyNumber","control","userProvidedThresholdValue", "userProvidedCqValue","runRecordedBy", "runDate","runTime","runPlatform","machineID","reactionConditions","reactionVolume","templateAmount","forwardPrimerBatch","reversePrimerBatch","dNTPConcentration", "primerConcentration","probeConcentration","Mg2+Concentration","polymeraseBatch","polymeraseConcentrations","thermocyclerParameters","pcrDataNotes","taxonID","establishmentMeans","assayName","assayOwnership","assayDescription","assayCitation","assayDate","geneTarget","geneSymbol","dilutions","replicates","primerR","primerF","probe","ampliconLength (bp)","probeFluorescentTag","dye(s)","quencher","probeModification","kingdom","phylum","class","order","family","genus","subgenus","species","vernacularName","organismScope","replicateID","extractName","analyst", "extractionDate", "extractionTime", "location", "extractionMethod", "methodCitation", "extractionNotes","tubePlateID","frozen", "fixed","dnaStorageLocation","extractMethodOfStorage","dnaVolume","quantificationMethod", "concentration(ng/ul)","stationID","collectorName","replicateName","collectionDate","collectionTime","storageID","DateOfStorage","methodOfStorage","minimumElevationInMeters","maximumElevationInMeters","verbatimElevation","minimumDepthInMeters","maximumDepthInMeters","verbatimDepth", "flowRate(m/s)", "filterType","filtrationDuration(mins)","volumeFiltered","processLocation","replicationNumber","riparianVegetationPercentageCover","dissolvedOxygen(mg/L)","waterTemperature(C)","pH","TSS(mg/L)","EC(uS/cm)","turbidity(NTU)","discharge","tide","chlorophyl","salinity(ppt)","contaminants(ng/g)","traceMetals(mg/kg)","organicContent(%)","microbialActivity","grainSize","replicateDataNotes","siteID","stationName","decimalLongitude","decimalLatitude","geographicRegionID","locality","estimatedPerimeter","estimatedSurfaceArea(m2)","siteType","siteLength(m2)","projectID","continent","country","stateProvince","municipality","projectCreationDate","projectName", "projectRecordedBy","projectOwner","projectContactEmail","projectDescription","InstitutionID","projectDataNotes","standardCurveID","standardCurveName","SCdate","SCrecordedBy","SCdataNotes")])
  
  return(metadata_table)
}

#Function removes null records (controls or empty runs) from associated fluorescence file based on if sample name contains 'null' in metadata file.
null_records_to_remove_flur <- function(meta_data, fluor_file) {
  if (length(which(grepl("null", tolower(meta_data$sampleName)))) != 0){
    fluor_file <- as.data.frame(fluor_file[-c(which(grepl("null", tolower(meta_data$sampleName)))+1)],)
    meta_data <- meta_data[-c(which(grepl("null", tolower(meta_data$sampleName)))), ]}
  scraped_files = list(fluor_file, meta_data)
  return (scraped_files)}

# merge raw fluorescence and metadata
merge_meta_flu_RDML <- function (processed_data_list){
  
  merged_file <- merge(processed_data_list[[2]], processed_data_list[[1]], by="wellLocation")
  merged_file <- merged_file %>% relocate(starts_with("well"), .after = "extractID")
  merged_file <- merged_file %>% relocate(starts_with("Cycle_"), .after = "control")
  
  return(merged_file)
}

# convert letters to numbers (reaction ID in RDML file must be numeric)
letter2number <- function(x) {utf8ToInt(x) - utf8ToInt("a") + 1L}

# create RDML file
process_RDML <- function(merged_file){
  fdata <- select(merged_file,contains("Cycle_Number"))
  fdata <- as.data.frame(t(fdata))
  fdata$cycle <- 1:nrow(fdata)
  fdata <- select(fdata,"cycle", everything())
  fdata <- sapply(fdata, as.numeric )
  
  fdata_descr <- data.frame(
    fdata.name = colnames(fdata)[2:ncol(fdata)],
    exp.id = as.character(merged_file$projectID),
    run.id = as.character(merged_file$runID),
    position= as.character(merged_file$wellLocation),
    #position=as.numeric(as.factor(merged_file$wellLocation)),
    #react.id = if(is.character(merged_file$assayID)) unname(sapply(tolower(merged_file$assayID), letter2number)) else (merged_file$assayID),
    react.id=if(is.character(merged_file$wellLocation)) unname(sapply(tolower(merged_file$wellLocation), letter2number)) else (merged_file$wellLocation),
    sample = as.character(merged_file$sampleName),
    sampleType=as.character(ifelse(is.na(merged_file$replicateName), "ctrl", merged_file$replicateName)),
    sample.type = as.character(replicate(nrow(merged_file),"unkn")),
    target = as.character(merged_file$geneTarget),
    target.dyeId = as.character(ifelse(is.na(merged_file$probeFluorescentTag), "Unknown",merged_file$probeFluorescentTag)),
    stringsAsFactors = FALSE
  )
  
  fdata_rdml <- RDML$new()
  # Add fluorescence data
  fdata_rdml$SetFData(fdata, fdata_descr)
  
  # Observe object
  fdata_rdml$AsDendrogram()
  return(fdata_rdml)
}
```

### Raw Data Processing 

```{r}
# Set File Names
fluorescence_data <- "MIC_experimental_raw_fluorescence_data.csv"
metadata_file <- "MIC_qPCR_Metadata_file.xlsx"

# Uncomment the lines below to run the biomeme version of the module
# fluorescence_data <-"biomemetwo3_experimental_raw_fluorescence_data.csv"
# metadata_file <- "biomemetwo3_qPCR_Metadata_file.xlsx"

# Platform Name
# "biomemefranklin", "MIC"
platform <- "MIC"

process_MIC_raw_data <- function(raw_fluorescence){
  raw_fluorescence <- as.data.frame(t(raw_fluorescence[, 2:ncol(raw_fluorescence)]))
  # add consistent column names
  #Changing column names to consistent with naming convention in database
  total_runs <- ncol(raw_fluorescence)
  colnames(raw_fluorescence) <- c(paste0("Cycle_Number", 1:total_runs))
  # add a column for merging with metadata
  raw_fluorescence$wellLocation <- as.numeric(gsub("[^0-9]", "", rownames(raw_fluorescence)))
  # move wellLocation to the front
  raw_fluorescence <- as.data.frame(raw_fluorescence[,c(ncol(raw_fluorescence),1:ncol(raw_fluorescence)-1)])
  print(paste0("number of wells: ", nrow(raw_fluorescence), ", number of cycles: ", ncol(raw_fluorescence)-1))
  return(raw_fluorescence)
}

process_biomeme_raw_data <- function(raw_fluorescence){
  #Creating dataframe with fluorescence values from biomem raw qPCR file
  end_row <-(which(grepl('Raw Fluorescence', raw_fluorescence$Run.Name))) - 2
  raw_fluorescence <- raw_fluorescence[11:end_row, 2:41]
  #Changing column names to consistent with naming convention in database
  total_runs <- ncol(raw_fluorescence)
  colnames(raw_fluorescence) <- c(paste0("Cycle_Number", 1:total_runs))
  #change all values to numeric
  cols = c(1:total_runs)
  raw_fluorescence[, cols] <- apply(raw_fluorescence[, cols], 2, function(x)  as.numeric(as.character(x)))
  # add wellLocation (key for merging)
  raw_fluorescence$wellLocation <- fluorescence_file[11:end_row, 1]
  raw_fluorescence <- as.data.frame(raw_fluorescence[,c(ncol(raw_fluorescence),1:ncol(raw_fluorescence)-1)])
  print(paste0("number of wells: ", nrow(raw_fluorescence), ", number of cycles: ", ncol(raw_fluorescence)-1))
  return (raw_fluorescence)
  }

if (platform=="MIC"){
  # process and format the MIC data
  processed_raw_flu <- process_MIC_raw_data(read.csv(fluorescence_data))
} else if (platform=="biomemefranklin"){
  # process and format the MIC data
  processed_raw_flu <- process_biomeme_raw_data(read.csv(fluorescence_data))
}

# ensure data type is numeric and convert if necessary
processed_raw_flu <- data_type_check(processed_raw_flu)

# process and format the metadata
metadata <- read_excel_allsheets(metadata_file)
formatted_metadata <- format_qPCR_metadata_RDML(metadata)

#remove control records
scraped_file_list <- null_records_to_remove_flur(formatted_metadata, processed_raw_flu)

# merge data files
merged_file <- merge_meta_flu_RDML(scraped_file_list)
# create RDML file
RDML.file <- process_RDML(merged_file)
# select a file name for export
outputname <- "/Users/abiyogasekaram/Downloads/6SP.rdml"
# export RDML file for later use
RDML.file$AsXML(outputname)

```


#### Loop Structure for converting multiple files into RDML

If there are multiple fluorescence files you would like to convert into RDML, the following loop operationalizes the conversion of files IF placed in the following directory/naming structure.

1) Place the fluorescence files and Metadata files into a directory.
2) Ensure a 1-to-1 naming structure between the fluorescence and metadata files (ex. Hanlon_Run1.csv and Hanlon_Run1.xlsx)
2) Replace the following flu_files and metadata_files code with the path to the directory.
3) Run the for loop. 
```{r}

file_directory_path <- "/Users/abiyogasekaram/Documents/MDMAPR/DataAnalysis2021/BrookTrout"
meta_files_path <- "/Users/abiyogasekaram/Documents/MDMAPR/DataAnalysis2021/BrookTrout/Metadata\ Files"
# vector of all fluorescence files
flu_files <- list.files(path = file_directory_path, pattern = ".csv", all.files = FALSE,full.names = T, recursive = FALSE,ignore.case = FALSE, include.dirs = FALSE, no.. = FALSE)

#vector of all the metadata files
metadata_files <- list.files(path = meta_files_path, pattern = ".xlsx", all.files = FALSE,full.names = T, recursive = FALSE,ignore.case = FALSE, include.dirs = FALSE, no.. = FALSE)

directory_path <- 

for (i in 1:length(flu_files)){
  fluorescence_data <- flu_files[i]
  metadata_file <- metadata_files[i]

  run_name <- sub(pattern = "(.*)\\..*$", replacement = "\\1", basename(flu_files[i]))

  # process and format the MIC data
  processed_raw_flu <- process_MIC_raw_data(read.csv(fluorescence_data))

  # ensure data type is numeric and convert if necessary
  processed_raw_flu <- data_type_check(processed_raw_flu)

  # process and format the metadata
  metadata <- read_excel_allsheets(metadata_file)
  formatted_metadata <- format_qPCR_metadata_RDML(metadata)

  #remove control records
  # scraped_file_list <- null_records_to_remove_flur(formatted_metadata, processed_raw_flu)
  scraped_file_list <- list(processed_raw_flu, formatted_metadata)

  # merge data files
  merged_file <- merge_meta_flu_RDML(scraped_file_list)
  # create RDML file
  RDML.file <- process_RDML(merged_file)
  # select a file name for export
  outputname <- paste(file_directory_path,run_name,".rdml", sep = "")
  # export RDML file for later use
  RDML.file$AsXML(outputname)
}
```


### Standard Curve Data Processing Module 

A separate RDML file will need to be created for the standard curve data. 

#### Functions for Standard Curve Data Only

```{r}
format_standardCurve_metadata <- function (standardCurve_metadata) {
  standardCurve_Table <- read_excel(standardCurve_metadata, sheet = 5)
  standardCurve_Table <- as.data.frame(standardCurve_Table[, c("SCresultID",
                                                                  "runID",
                                                                  "pcrChemistryID",
                                                                  "standardCurveID",
                                                                  "wellLocation",
                                                                  "sampleName",
                                                                  "copyNumber",
                                                                  "control",
                                                                  "standardConc",
                                                                  "userProvidedThresholdValue",
                                                                  "userProvidedCqValue",
                                                                  "runRecordedBy",
                                                                  "runDate",
                                                                  "runTime",
                                                                  "runPlatform",
                                                                  "machineID",
                                                                  "reactionConditions",
                                                                  "reactionVolume",
                                                                  "templateAmount",
                                                                  "forwardPrimerBatch",
                                                                  "reversePrimerBatch",
                                                                  "dNTPConcentration",
                                                                  "primerConcentration",
                                                                  "probeConcentration",
                                                                  "Mg2+Concentration",
                                                                  "polymeraseBatch",
                                                                  "polymeraseConcentrations",
                                                                  "thermocyclerParameters",
                                                                  "pcrDataNotes",
                                                                  "assayID",
                                                                  "standardCurveName",
                                                                  "SCdate",
                                                                  "SCrecordedBy",
                                                                  "SCdataNotes",
                                                                  "LOD",
                                                                  "LOQ")])
  return (standardCurve_Table)
}

merge_meta_flu_std_RDML <- function (processed_data_list){
  merged_file <- merge(processed_data_list[[2]], processed_data_list[[1]], by="wellLocation")
  merged_file <- merged_file %>% relocate(starts_with("Cycle_"), .after = "control")
  return(merged_file)
}

```



#### RDML Processing for Std. Curve
```{r}
process_RDML_Std <- function(merged_file){
  fdata <- select(merged_file,contains("Cycle_Number"))
  fdata <- as.data.frame(t(fdata))
  fdata$cycle <- 1:nrow(fdata)
  fdata <- select(fdata,"cycle", everything())
  fdata <- sapply(fdata, as.numeric)
  fdata <- as.data.frame(fdata)
  
  fdata_descr <- data.frame(
    fdata.name = colnames(fdata)[2:ncol(fdata)],
    exp.id = merged_file$standardCurveID,
    run.id = merged_file$runID,
    position=merged_file$wellLocation,
    #position=as.numeric(as.factor(merged_file$wellLocation)),
    #react.id = if(is.character(merged_file$assayID)) unname(sapply(tolower(merged_file$assayID), letter2number)) else (merged_file$assayID),
    react.id = if(is.character(merged_file$wellLocation)) unname(sapply(tolower(merged_file$wellLocation), letter2number)) else (merged_file$wellLocation),
    sample = replicate(nrow(merged_file),"Standard Curve"),
    #sampleType=merged_file$replicateName,
    sample.type = replicate(nrow(merged_file),"std"),
    target = replicate(nrow(merged_file),"unknown"),
    target.dyeId = ifelse(is.na(merged_file$pcrChemistryID), "Unknown",merged_file$pcrChemistryID),
    stringsAsFactors = FALSE
  )
  
  fdata_rdml <- RDML$new()
  # Add fluorescence data
  fdata_rdml$SetFData(fdata, fdata_descr)
  
  # Observe object
  fdata_rdml$AsDendrogram()
  return(fdata_rdml)
}
```

### Std. Curve RDML Creation

set variables for platform, path to fluorescence data, path to metadata template file
```{r}
platform <- "MIC"

std_fluorescence_data <- "MIC_standard_curve_raw_fluorescence_data.csv"
metadata_file <- "MIC_qPCR_Metadata_file.xlsx"

```


```{r}

# same raw data processing function can be used as above
if (platform=="MIC"){
  # process and format the MIC data
  std_fluorescence <- process_MIC_raw_data(read.csv(std_fluorescence_data))
} else if (platform=="biomemefranklin"){
  # process and format the MIC data
  std_fluorescence <- process_biomeme_raw_data(read.csv(std_fluorescence_data))
}


# ensure data type is numeric and convert if necessary
std_fluorescence <- data_type_check(std_fluorescence)

# process and format the metadata
formatted_metadata <- format_standardCurve_metadata(metadata_file)

#remove control records
scraped_file_list <- null_records_to_remove_flur(formatted_metadata, std_fluorescence)

# merge data files
merged_file <- merge_meta_flu_std_RDML(scraped_file_list)
# create RDML file
RDML.std.file <- process_RDML_Std(merged_file)
# select a file name for export
outputname <- "processed_rdml_std_file.RDML"
# export RDML file for later use
RDML.std.file$AsXML(outputname)

```
